Formalizing the Edmonds-Karp Algorithm

### 2016-03-02 paper/root.tex
eval-section, started concl

  \subsection{Manipulating Residual Graphs Directly}    
  In a next step, we observe that the algorithm is phrased in terms of a flow which is updated until it is maximal. However,
  in order to update the flow, the residual graph is used, which is a combination of the current flow and the network.
  Obviously, computing the complete residual graph each time before searching for an augmenting path would be a bad idea. 
  One solution to this problem is to compute the edges of the residual graph on demand from the network and the current flow.
  Although this solution seems to be common, it has the disadvantage that for each edge of the residual graph, two (or even three) 
  edges of the network and the flow have to be accessed. As edges of the residual graph are accessed in the inner loop, during 
  the BFS, these operations are time critical.

@@

  \subsection{Using Efficient Data Structures}    
  In a final step, we have to chose efficient data structures for the algorithm. The objects that the algorithm works with are
  capacities, nodes, edges, the graph of the input network, residual graphs, the adjacency map, paths, and the resulting maximum flow.

@@

\section{Benchmarking}
  What implementations did we compare: Authors? Where do implementations come from? 
  [We must convince the reader that we did not intentionally chose bad implementations to compare us against!]

  Comparsion of algorithms (Modified data structure: Flow vs Flow-like vs ResGraph)
    Computing successors in BFS: Filter by visited set first.
      -> Technically more challenging, when computing on flow: 1 access first vs 2 accesses first. On resGraph: No advantage in memory accesses (However, plain array access vs. matrix access)
    Computing bottleneck during BFS. Saves one iteration over the path, at the cost of one extra memory access per discovered node.
    For our benchmarks, we observed that the BFS discovers considerably more nodes than the length of the ultimately returned path, 
    and the overall code was faster with the extra iteration for bottleneck computation.

\section{Conclusion} 
  ... and related work
    Mizar-Formalization: What did they do.

>>>>

  \subsection{Manipulating Residual Graphs Directly}\label{sec:impl_res_graph}
  In a next step, we observe that the algorithm is phrased in terms of a flow which is updated until it is maximal. However,
  in order to update the flow, the residual graph is used, which is a combination of the current flow and the network.
  Obviously, computing the complete residual graph each time before searching for an augmenting path would be a bad idea. 
  One solution to this problem is to compute the edges of the residual graph on demand from the network and the current flow.
  Although this solution seems to be common, it has the disadvantage that for each edge of the residual graph, two (or even three) 
  edges of the network and the flow have to be accessed. As edges of the residual graph are accessed in the inner loop, during 
  the BFS, these operations are time critical.

@@

  \subsection{Using Efficient Data Structures}\label{sec:impl_datastructures}
  In a final step, we have to chose efficient data structures for the algorithm. The objects that the algorithm works with are
  capacities, nodes, edges, the graph of the input network, residual graphs, the adjacency map, paths, and the resulting maximum flow.

@@

\section{Benchmarking}
  We have compared the running time of our algorithm against an unverified reference implementation in Java, taken from Sedgewick and Wayne's book on algorithms~\cite{SeWa11}.
  We have done the comparison on randomly generated sparse and dense networks, the sparse networks having a \emph{density} ($\frac{|E|}{|V|(|V| - 1)}$) of $0.02$, and the dense networks having a density of $0.25$. For sparse networks, we varied the number of nodes between $1000$ and $5000$, for dense networks between $1000$ and $1400$.
  The results are shown in Table~\ref{tab:benchmark}. 
  \begin{wrapfigure}{o}{.4\textwidth}
  \vspace{-20pt}
  \begin{center}
  \begin{tabular}{|c|c|c|}
  \hline
  Name        & SML       & Java   \\
  \hline\hline
  1000-sparse & 204       & 15     \\
  2000-sparse & 1430      & 231    \\
  3000-sparse & 4928      & 828    \\
  4000-sparse & 9299      & 2145   \\
  5000-sparse & 23625     & 4868   \\
  1000-dense  & 6575      & 4011   \\
  1100-dense  & 8067      & 5487   \\
  1200-dense  & 10892     & 7011   \\
  1300-dense  & 14070     & 9815   \\
  1400-dense  & 17317     & 12156  \\
  \hline
  \end{tabular}
  \end{center}
  \vspace{-15pt}
  \caption{Dijkstra benchmark}\label{tab:benchmark}
  \vspace{-20pt}
  \end{wrapfigure}

  We observe that, for sparse graphs, the Java implementation is roughly faster by a factor of 6, while for dense graphs, our implementation comes quite close 
  to the Java implementation. We have no full explanation for this phenomenon. Note that the Java implementation operates on flows, while our implementation 
  operates on residual graphs (\cf Section~\ref{sec:impl_res_graph}). Moreover, the Java implementation does not store the augmenting 
  path in an intermediate list, but uses the predecessor map computed by the BFS directly (\cf Section~\ref{sec:impl_datastructures}).

  Finally note that a carefully optimized C++ implementation of the algorithm is only slightly faster than the Java implementation for sparse graphs,
  but roughly one order of magnitude faster for dense graphs. We leave it to future work to investigate this issue, and conclude that we were able to produce
  a reasonably fast verified implementation, without doing any low-level optimizations, but relying on the code generated by the Isabelle/HOL code generator.

% 
%   What implementations did we compare: Authors? Where do implementations come from? 
%   [We must convince the reader that we did not intentionally chose bad implementations to compare us against!]
% 
%   Comparsion of algorithms (Modified data structure: Flow vs Flow-like vs ResGraph)
%     Computing successors in BFS: Filter by visited set first.
%       -> Technically more challenging, when computing on flow: 1 access first vs 2 accesses first. On resGraph: No advantage in memory accesses (However, plain array access vs. matrix access)
%     Computing bottleneck during BFS. Saves one iteration over the path, at the cost of one extra memory access per discovered node.
%     For our benchmarks, we observed that the BFS discovers considerably more nodes than the length of the ultimately returned path, 
%     and the overall code was faster with the extra iteration for bottleneck computation.

\section{Conclusion}
  In this paper, we have presented a verification of the Ford-Fulkerson method, using a stepwise refinement approach.
  Starting with a proof of the Ford-Fulkerson theorem, we have verified the generic Ford-Fulkerson method, then 
  specialized it to the Edmonds-Karp algorithm and proved the upper bound $O(|E||V|)$ for the number of outer loop iterations.
  We then used several refinement steps to derive an efficiently executable implementation of the algorithm, 
  involving a verified breadth first search to obtain shortest augmenting paths. 
  Finally, we added a verified algorithm to check whether the input is a valid network, and generated executable code in SML.
  Our implementation compares well to a reference implementation in Java.

  Our formalization has combined several techniques to achieve an elegant and accessible formalization: The Isar-proof language~\cite{}
  allows us to write formal proofs that closely follow informal proofs found in textbooks. This way, we have created a completely rigorous, but 
  still accessible, proof of the Ford-Fulkerson theorem. The Isabelle Refinement Framework allows us to first formalize an algorithm abstractly, 
  and then refine it, in a modular fashion, towards an efficient implementation. This allowed us to present the Ford-Fulkerson method on a level 
  of abstraction that closely resembles pseudocode presentations found in textbooks, and then formally link this presentation to an efficient
  implementation. Moreover, modularity of refinement allowed us to develop the breadth first search algorithm independently, and later link it to the 
  main algorithm. The BFS algorithm can be reused as building block for other algorithms. Also the data structures are re-usable, and in fact, the 
  only new data structure we had to implement where the adjacency matrices. The other data structures where readily available in the growing library 
  of the Sepref tool~\cite{La15} and the Isabelle Collection Framework~\cite{LaTu12}.

  \subsection{Related Work}
  We are only aware of one other formalization of the Ford-Fulkerson method conducted in Mizar~\cite{}. 
  It formalizes a fixed instance of the Ford Fulkerson method on a rather abstract level. No code has been generated, 
  nor does it contain a complexity analysis.

  Apart from our own work~\cite{La14,NoLa12}, there are several other verifications of graph algorithms, using different techniques. 
  Top-Down
  Bottom-Up: Noschinski verifies a checker for (non-) planarity certificates using a bottom-up approach. Starting at a C implementation,
  the AutoCorres tool generates a monadic representation of the program in Isabelle. 
  XXX, ctd here

  Unfortunately, it

  ... and related work
    Mizar-Formalization: What did they do.
